---
title: "SL_assignment6_Ian"
author: "Ian Wallgren"
date: "2023-03-24"
output: pdf_document
---


Consider the nonparametric regression of cnt as a function of instant.
Estimate the regression function m(instant) of cnt as a function of
instant using a cubic regression splines estimated with the R function
smooth.splines and choosing the smoothing parameter by Generalized
Cross Validation.

```{r}

x <- bikes$instant
y <- bikes$cnt


library(interp)
#install.packages("mblm")
library(mblm)


#choosing the optimal smoothing param by setting cv to TRUE
cubic_spline_fit = smooth.spline(x=x, y=y, cv = TRUE)


#a)
cat("The value of the chosen penalty term amplifier lambda is: ",cubic_spline_fit$lambda)

#b)
cat("The corresponding number of degrees of freedom is: ",cubic_spline_fit$df)

#c)
#we take minus to since the first and last one does not "count" as knots
num_knots = cubic_spline_fit$fit$nk - -2
cat("Number of knots used is: ",num_knots)

#d)
plot(x, y, xlab = "instant", ylab = "cnt")
lines(cubic_spline_fit, col = "dark green", lwd = 2)
legend("topleft", legend = "Fitted cubic spline", col = "dark green", lwd = 2)
title("Scatterplot of cnt vs. instant with cubic spline fitted line")


```

Define a new variable cnt.5000 taking the value 1 for days such that the
number of total rental bikes is larger than or equal to 5000, on 0 otherwise.

```{r}

#creating an empty col
bikes$cnt_5000 <- rep(NA, dim(bikes)[[1]])

#there should be a fancier way of doing this, not used to R hehe
for (x in 1:dim(bikes)[[1]]){
  if (bikes$cnt[x] > 5000){
    bikes$cnt_5000[x] = 1 
  } else {
    bikes$cnt_5000[x] = 0
  }
  
}



```

a) Use the function logistic.IRWLS.splines to fit the non-parametric
binary regression cnt.5000 as a function of the temperature, using
df=6.

```{r}

#copy paste function
logistic.IRWLS.splines <- function(x,y,weights.out=1,x.new=x,
                           df=6,spar=NULL, 
                           all.knots = FALSE, nknots = .nknots.smspl,  
                           max.iter=10,eps.beta=1e-5,
                           plts=FALSE){
  if (plts&(dim(as.matrix(x))[2]>1)){
    plts<-FALSE
    warning("Plots are valid only when dim(x)[2]==1")
  }
  # Step 0
  stop.rule <- FALSE
  iter <- 0
  theta.0 <- fitted(lm(y~x)) 
  
  while (!stop.rule){
    iter <- iter + 1 
    # step 1
    p <- p.from.theta(theta.0)
    ps.e <- (y-p)/(p*(1-p))
    z <- theta.0 + ps.e 
    wt <- p*(1-p) *weights.out
    
    if (plts){
      op<-par(mfrow=c(1,2))
      plot(x,y,cex=8*wt)
      lines(x,p,col=2)
      plot(x,z,cex=8*wt)
      lines(x,theta.0,col=2)
      par(op)
    }
    
    spline.1 <- smooth.spline(x,z,w=wt,df=df,spar=spar,
                              all.knots = all.knots,
                              nknots = nknots) 
    theta.1 <- predict(spline.1,x=x)$y
    
    # checking stop rules
    if ((iter>=max.iter)|(sum((theta.1-theta.0)^2)<eps.beta)){
      stop.rule<-TRUE
    } else {
      theta.0 <- theta.1
    }
  }
  
  p <- p.from.theta(theta.1)
  resid.devi <- sum(-2*dbinom(y,1,p,log=TRUE))
  
  return(list(fitted.values=p,
              theta.x=theta.1,
              df=spline.1$df,
              predicted.values=p.from.theta(predict(spline.1,x=x.new)$y),
              residual.deviance=resid.devi)
  )
} 

# This is an auxiliary function for logistic.IRWLS.splines
#
# p.from.theta(theta,x) 
# computes the vector of probabilities p, with elements
# p_i=\hat{Pr}(Y_i=1|X=x_i), i=1,...,n,
# where x_i is the ith row of matrix "x",
# a logistic link is assumed, and "theta" is the vector 
# with the nonparametric estimations of 
# log( Pr(Y_i=1|X=x_i)/(1-Pr(Y_i=1|X=x_i)) ), i=1,...,n.
p.from.theta <- function(theta.x){
  p.x <- 1/(1+exp(-theta.x))
  return(p.x)
}


```


Applying the function to the obtain desired regression

In which range of temperatures is
Pr(cnt >= 5000|temp) larger than 0,5?

```{r}

log_reg_tmp_cnt5000 = logistic.IRWLS.splines(bikes$temp,bikes$cnt_5000,df=6)

plot(log_reg_tmp_cnt5000$predicted.values)

#finding indices where Pr(cnt >= 5000 | temp) > 0.5
idx <- which(log_reg_tmp_cnt5000$predicted.values > 0.5)

plot(bikes$temp[idx])

min_temp = min(bikes$temp[idx])
max_temp = max(bikes$temp[idx])
range_temp = c(min_temp,max_temp)

cat("The range of the temperature where the probability of cnt_5000 to be greater than 0.5 is: ",range_temp)
```

Choose the parameter df by k-fold log-likelihood cross validation
with k = 5 and using df.v = 3:15 as the set of possible values for
df.

```{r}

df_v = seq(3,15)




# modifying Pedro's function to k-fold instead of leave one out
loglik.KCV <- function(x,y,df){
  df_v_res = numeric(15-3)
  n <- length(x)
  n_div_k = round(n / 5)
  
  pred <- sapply(1:n_div_k, 
      function(k,x,y,df){
        
        logistic.IRWLS.splines(x=bikes$temp[-k],y=bikes$cnt_5000[-k],df=df)$predicted.values
        
         #sm.binomial(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
      },   x,y,df)
  return(-sum( y*log(pred/(1-pred)) + log(1-pred) )/n)
}

degfree_kfold = function(x,y,dfv,method=loglik.KCV){
  i = 0
  for (df in dfv){
    i = i+1
    df_v_res[i] = method(x,y,df) 
  }
  return(list(df       = df_v,
              df_v_res = df_v_res,
              best_df  = df_v[which.min(df_v_res)]))
  
  
  
}

degfree_kfold(x=bikes$temp,y=bikes$cnt_5000,df_v)

#best degree of freedom choice turns out to be 3
```







